---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
```


```{r}

## Retinalintensity = surfaceshade / lightintensity
retinal.intensity <- 0.2

## If retinal intensity remains constant go through all combinations
## of light intensity which could have produced this retinal intensity
light.intensity.space <- seq(0.01, 1, 0.01)

## Surface intensity space 
surface.intensity.space <- retinal.intensity / light.intensity.space

## Plot the light intensity vs surface intensity Likelihood space 
## This is the correct plot: checked the answers 
plot(light.intensity.space, surface.intensity.space, type = "l",
     xlab = "Light Intensity", ylab = "Surface Intensity",
     xlim = c(0, 1), ylim = c(0,1))

## Strong prior that the light intensity was between 0.2 and 0.4 and nothing else 
## Draw this on a plot - should be a uniform distribution between 0.2 and 0.4

```


```{r}

stimulus <- seq(from = -10, to = 10, by = 0.01)

## Prior distribution - get the density of a gaussian (normal) distribution 
prior <- dnorm(stimulus, mean = 0, sd = 2)
prior <- prior /sum(prior)

xt <- 3.2 # observation 

## Likelihood is the uncertianty around the measurement, x(t)
## Probability density at a particular value of xt (holding the observation constant)
## and getting the probability for each stimulus value, under the hypothesis of mean, xt
likelihood <- dnorm(stimulus, mean = xt, sd = 1)
likelihood <- likelihood / sum(likelihood)

## Posterior 
posterior <- prior * likelihood
posterior <- posterior / sum(posterior)
MAP <- max(posterior)

## Plot the prior distribution 
plot(stimulus, prior, 
     type = "l", 
     ylab = "Probability", xlab = "Stimulus Values",
     main =  "Prior Distribution")

## Plot the likelihood function 
plot(stimulus, likelihood, 
     type = "l", 
     ylab = "Probability", xlab = "Stimulus Values",
     main =  "Likelihood Function")
abline(v =  xt, col = "red")

## Plot the posterior distribution 
plot(stimulus, posterior,
     type = "l", 
     ylab = "Probability", xlab = "Stimulus Values",
     main =  "Posterior Distribution")
abline(v =  MAP, col = "red")

## Plot them all on the same graph - notice how the likelihood is pulled
## by the prior, and that's where the posterior ends up 
plot(stimulus, prior, 
     type = "l", 
     ylab = "Probability", xlab = "Stimulus Values",
     col = "red",
     main =  "", ylim = c(0, 0.0065))
lines(stimulus, likelihood, col = "blue")
lines(stimulus, posterior, col = "black")
legend("topright", legend = c("Prior", "Likelihood", "Posterior"),
       col = c("red", "blue", "black"), lty = 1:3, cex = 0.8,
       box.col = NA)

  
```

Change point tutorial - this is hard!! 
```{r}

## Setup experiment parameters 
s <- c(-1, -1, -1, -1, -1, 1, 1, 1, 1, 1) # change point occurs at 5-6 boundary 

sigma <- 1 ## change later 

## These are the noisy observations 
#xt <- rnorm(10, mean = s, sd = 1)
xt <- c(-0.46, 0.83, -3.26, -0.14, -0.68, -2.31, 0.57, 1.34, 4.58, 3.77) ## used in the paper 

# The trials that we are iterating over 
t.change <- seq(1, 10)

## Plot values of the stimulus along with each iteration of change points 
plot(t.change, xt, type = "l", xlab = "Time", ylab = "measurement")
abline(a = 0, b = 0, col = "red")

## Posterior probability calculation 
posterior.ch <- rep(NA, length(xt))

## Posterior calculation we are iterating over possible change point hypotheses 
## so we are evaluating whether a change point occured at 1:T. 
for (t in 1:length(s)){
  posterior.ch[t] <- exp( 2  *  sum( xt[t:length(s)] ) ) 
}

## Normalise posterior so that it is between 0 and 1 by dividing by the sum 
posterior.ch <- posterior.ch/sum(posterior.ch)

## Stimulius category 
plot(1:length(xt), s,
     xlab = "Trial Number", ylab = "Stimulus Category (s) (1 = A, -1 = B)",
     type = "l")

## Noisy representation of stimulus category 
plot(1:length(xt), xt,
      xlab = "Trial Number", ylab = "Noisy measurement of stimulus x(t)",
     type = "l")

## Plot posterior probability of a change point at each theoretical value of tchange 
plot(1:length(xt), posterior.ch, 
     xlab = "Possiblity that a change point occured on trial t", ylab = "Posterior",
     type = "l")

```


Assume s = 1 and T = 10. Vary the true change point tchange from 1 to T. For each value of tchange, we simulate 10,000 trials (or more if possible). On each simulated trial,
Based on tchange; specify the stimulus sequence s.
Simulate a measurement sequence x from s.
Apply the decision rule to each measurement sequence. The output is the simulated observer’s response, namely an estimate of tchange.
Determine whether the response was correct.

## Todo --- 
## Implement decision rule - maximum of posterior 
## Determine response (correct vs incorrect) - compare whether the change point was on that trial 
## Get proportion correct for each true value of "tchange"
## Vary sigma (1,23) and T from 2 - 16 in steps of 2 - plot overall prop correct as a function of T 
```{r}

sim.data <- function(tchange, sigma, time.scale){
  ## ---------------------------------------------------------------------------------------
  ## Description: this function simulates a dataset for a specific value of the following inputs:
  ## Inputs: 
  ## - tchange = the trial on which a change point occurs (numerical integer)
  ## - sigma = the noise you assume in the measurement (SD) (numerical)
  ## - time.scale = the number of trials. If this number is 10, then trials go from 1:10. (numerical)
  ## Ouput: 
  ## - data = simulated dataset in the form of a data frame which contains the inputs as columns and 
  ## the posterior probability of a change point, given the simulated data 
  ## ---------------------------------------------------------------------------------------
  
  data <- data.frame(
    time = seq(1, time.scale),
    change.point = rep(tchange, time.scale),
    stim = c(rep(-1, length(1:tchange)), rep(1, length((tchange+1) : time.scale)))
  )
  
  ## Generate noisy observations 
  data$obs <- rnorm(time.scale, mean = data$stim, sd = sigma)
  
  ## Generate posterior probability 
  data$post.prob <- NA
  
  ## Posterior probability updating - each iteration of this loop is saying: "what is the posterior probabiliy
  ## that a change point occured on trial t, if I look at the observations from t = the value in the loop to 
  ## the last trial I saw 
  for (t in 1:time.scale){
    data$post.prob[t] <- exp(2 * sum(data$obs[t:time.scale]) )
  }
  
  ## Perform normalisation by dividing by the sum 
  data$post.prob <- data$post.prob / sum(data$post.prob)
  
  ## Convert to tibble for fun. 
  data <- as_tibble(data)

  return(data)
}


## todo(CVH): insert parameters here 

## Created nested datasets to feed into the simulation 
sim <- as_tibble(expand.grid(
  tchange = 1:9, ## the trial on which there is a change point 
  sigma = 1:3 ## noise parameter
))

## Run function 
sim <- sim %>% 
  mutate(
    data = map(tchange, sim.data, sigma, 10)
  ) 
sim$simNo <- as.factor(1:dim(sim)[1])

## At the moment we have data nested in a line of a tibble (for matlab users, this is like a cell array)
## We want to get rid of this nesting structure, so we will perform an unnesting of it 
## If you want to have a look at before vs after, just type "sim" in the console before you run this line, and after 
sim <- sim %>%
  unnest(data) 

## Note: posterior probability will be same for trials in each simNo because it's the max 
## of ALL of the post.prob values for a single simulation - yeah, it's redundant, but who cares, it means 
## it fits in the same data frame so whatever! 

## MAP stands for maximum-a-posteriori and is just the highest value of the posterior; that is, the value of 
## tchange with the highest probability. We would expect that given we simulated the data with change points on 
## particular trials, the MAP for each dataset would occur around the value of tchange that we simulated the change 
## point to be from. I.e. if a change point occured at trial 5, the posterior should peak (be at a maximum) around trial 5.
sim <- sim %>% 
  group_by(simNo) %>%
  mutate(
    MAP = max(post.prob), ## maximum value of posterior (MAP estimate)
    pred.change.point = which(post.prob == MAP)-1, ## figure out which trial the MAP occured on (infer where the model thinks a change point occured)
  )

## TODO(CVH): you were up to here 
## If the model and the actual change point where we simulated match, the values of change.point and pred.change.point should match
sim$correct <- sim$change.point == sim$pred.change.point

## Calculate proportion correct 
sim %>% 
  group_by(change.point, sigma) %>% 
  mutate(
    prop.correct = length(sim$correct)
  )

sim$simNo 
```

```{r}

## ------- Plot the posterior probability over change points -------------
## Below, each facet of the plot is the trial we simulated a change point over 
## If you would prefer a smoothed line, you can comment the geom_line() line and 
## uncomment the geom_smooth() line. I have used some colourblind palettes below, 
## but keep in mind R libraries for colourblind palettes are also available (google)

# http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/ (this is where I got these palettes from)
## Palette with grey 
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
  
## Plot simulations for each noise level 
sim %>% 
  ggplot(., aes(x = time, y = post.prob, group = as.factor(simNo))) + 
  geom_line(aes(x = time, y = post.prob, group = simNo, colour = as.factor(sigma))) + 
  #geom_smooth(aes(y = post.prob, group = simNo, colour = as.factor(sigma)), se = FALSE, size = 0.5) +
  facet_wrap(~change.point) +  ## we want a separate panel for change point we generated data from 
  ## Legend settings: if you want to read more about legends: https://ggplot2-book.org/guides.html (Danielle Navarro wrote this)
  scale_colour_manual(
    name = "Noise",
    values = cbbPalette[2:4], 
  ) + 
  xlab("Trial Number") + ylab("Posterior Probability \n of a Change Point") 
  

## Plot the stimulus distribution we simulated from - you will see exactly when the change point occured here
sim %>% 
  ggplot(., aes(x = time, y = stim, group = as.factor(simNo))) + 
  geom_line(aes(x = time, y = stim, group = simNo, colour = as.factor(sigma))) + 
  #geom_smooth(aes(y = post.prob, group = simNo, colour = as.factor(sigma)), se = FALSE, size = 0.5) +
  facet_wrap(~change.point) +  ## we want a separate panel for change point we generated data from 
  ## Legend settings: if you want to read more about legends: https://ggplot2-book.org/guides.html (Danielle Navarro wrote this)
  scale_colour_manual(
    name = "Noise",
    values = cbbPalette[2:4], 
  ) + 
  xlab("Trial Number") + ylab("Stimulus (+ = Category A, - = Stimulus B)") 
  
## It might also be cool to plot the observed sequence, you can then eyeball it to sanity check
## You should hopefully observe that when the distribution changes, so does the posterior estimate of 
## the change point trial - however, given we are drawing from a normal distribution, this is not trivial! 
## If you look at the plot you can see it's tricky to know when a change point occured. 
sim %>% 
  ggplot(., aes(x = time, y = obs, group = as.factor(simNo))) + 
  geom_line(aes(x = time, y = obs, group = simNo, colour = as.factor(sigma))) + 
  #geom_smooth(aes(y = post.prob, group = simNo, colour = as.factor(sigma)), se = FALSE, size = 0.5) +
  facet_wrap(~change.point) +  ## we want a separate panel for change point we generated data from 
  ## Legend settings: if you want to read more about legends: https://ggplot2-book.org/guides.html (Danielle Navarro wrote this)
  scale_colour_manual(
    name = "Noise",
    values = cbbPalette[2:4], 
  ) + 
  xlab("Trial Number") + ylab("Observations (+ = Category A, - = Stimulus B)") 
         
```

